---
title: "Advanced Topics"
format: html
editor: visual
---

TODO:

* everything
* update solutions to match new data [to be done Wed!]

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(30790)
```

## Advanced topics in Stratigraphic Paleobiology

In this practical, we go through some advanced topics, such as modeling taphonomic effects, and transforming phylogenetic data (time trees and fossils)

## Introduction

TBA

## Code cheat sheet

### Modeling taphonomy

Modeling taphonomy follows the same principles as modeling ecology. You need to define two function:

* a function `ctc` describing the change in taphonomic conditions with time (corresponding to the change in gradient with time for ecology)

* a function `pres_potential` that specifies how likely a fossil is to be preserved given some taphonomic conditions (equivalent to the niche definiton for ecology)

Both can then be passed to the function `apply_taphonomy` (instead of `apply_ecology`)

```{r}
# insert example here
```

Just as ecology, this can also be used for trait evolution

```{r}
# instert example with paleoTS here
```


## Tasks

::: {.callout-important}
### Task 1: Taphonomic effects

Model a taxon that has a preservation probability of 0.5 in water depths lower than 20 m and a preservation probablity of 0.9 in water deeper than 20 m.

How abundant is this taxon at different locations on the platform?
From your modeling results along, is there any way to distinguish these taphonomic effect from an ecological effect?

:::

::: {.callout-important}
### Task 2: Phylogenetic trees in stratigraphic context

Simulate an FBD (time)tree over the duration of the age-depth model and transform it at different locations using the age-depth model.

Simulate some fossils along the tree using the `FossilSim` package. 

How are the trees affected at the different locations? What are the most obvious visual changes in their topology?

:::


## Solutions

::: {.callout-note collapse="true"}
### Solution Task 1: Taphonomic effects

:::

::: {.callout-note collapse="true"}
### Solution Task 2: Phylogenetic trees in stratigraphic context

:::
