---
title: "Getting Started"
format: html
editor: visual
bibliography: refs/refs.bib
---

## Overview

This document describes how to set up your machine so you can use the workshop materials

## Requirements

The workshop materials requires R version 4.2 or later, which can be downloaded on [CRAN](https://cran.r-project.org/). You can use R from the terminal or use an IDE of your choice, e.g., [Rstudio](https://posit.co/download/rstudio-desktop/).

The workshop assumes basic knowledge of R, meaning you should be able to write short code snippets and execute them. However, the focus of the workshop is the scientific content, so we provide code templates and snippets for you to work through. If you're unsure of your R level you can work through [this course](https://swcarpentry.github.io/r-novice-inflammation/) by the software carpentry, which provides a good bases.

## Package installation

The workshop relies on the `StratPal` and `admtools` packages @hohmann_stratpal_mee_2025 @hohmann_admtools_gchron_2025 , which you can install from CRAN using the command

```{r, eval=FALSE}
install.packages("StratPal")
install.packages("admtools")
```

This also installs all dependencies, e.g., the `FossilSim` package for simulating fossils and the `paleoTS` package for analysis of fossil time series @barido-sottani2019 @hunt2006.

The two packages have different responsibilities:

-   `admtools` takes care of the *stratigraphy* components, e.g., work with age-depth models and incompleteness

-   `StratPal` takes care of the `Paleobiology` components, e.g., trait simulations, niche modeling, and taphonomy, and provides example stratigraphic architectures.

## Piping

We will make extensive use of the *piping operator* `|>` in this workshop, which is a feature of base R since version 4.2.

You can read `|>` as "take what is on the left side of `|>` and use it as first argument on the right side of `|>`". For example, to plot a histogram of normally distributed random variables, you can use

```{r}
rnorm(100) |>    # 100 normally distrubuted random variables                      
  hist(xlab = "x",
       main = "Normal distribution")
```

You can chain as many pipe operators together as you like:

```{r}
100 |>
rnorm() |>    # 100 normally distrubuted random variables                      
  hist(xlab = "x",
       main = "Normal distribution")
```

This solves two problems:

-   There is no need to assign intermediate variables that will then clutter your workspace. For example, to calculate the mean of the square root of uniformly distributed random numbers, it is tempting to do

    ```{r}
    t1 = runif(1000)
    t2 = sqrt(t1)
    x = mean(t2) # creates 2 unnecessary variables t2 and t2
    x = runif(1000) |> sqrt() |> mean() # only the intended result is saved
    ```

<!-- -->

-   You can read the code from left to right, making it easier to comprehend as the order of execution follows the order of reading. Continuing the above, it is tempting to write

    ```{r}
    x = mean(sqrt(runif(1000))) # has to be read from "inside out" (right to left)
    # instead of left to right, following the order of execution
    x = runif(1000) |> sqrt() |> mean()
    ```


