---
title: "Trait Evolution"
format: html
editor: visual
bibliography: refs/refs.bib
---

ToDo:

-   write Intro
-   write solution for task 4

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(15487)
```

## Trait evolution in stratigraphic context

This practical explores how trait evolution is influenced by stratigraphic architectures.

## Introduction

## Coding cheat sheet

### Modes of evolution in the time domain

There are three different modes of evolution implemented in the `StratPal` package: `stasis`, `random_walk`, and `ornstein_uhlenbeck`. You can plot them using piping via

```{r}
library(StratPal)
library(admtools)
pos = "2km"
adm = tp_to_adm(t = scenarioA$t_myr,        
                             h = scenarioA$h_m[,pos],
                             T_unit = "Myr",
                             L_unit = "m")
temp_res = 0.001                                              # temporal resolution in Myr
seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk(sigma = 1, mu = 0, y0 = 0) |>                   # simulate trait evolution
  plot(xlab = paste0("Time [",get_T_unit(adm),"]"),           # plot
       ylab = "Trait value",
       type = "l") 
  

```

To plot other modes, replace `random_walk` by one of the other modes (`stasis` or `ornstein_uhlenbeck`).

### Transforming trait evolution into the stratigraphic domain

To transform trait evolution into the stratigraphic domain, insert `time_to_strat` with the age-depth model of your choice into your pipeline:

```{r}
temp_res = 0.001                                              # temporal resolution in Myr
seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk(sigma = 1, mu = 3, y0 = 0) |>                   # simulate trait evolution
  time_to_strat(adm, destructive = FALSE) |>                  # transform into stratigraphic domain
  plot(xlab = paste0("Stratigraphic position [",get_L_unit(adm),"]"), # plot
       ylab = "Trait value",
       type = "l",
       orientation = "lr") 
```

This will transform your simulations from the time domain into the stratigraphic domain.

### Model selection using the paleoTS package

To work with model selection, we need to simulate traits of individual specimems, not just means of populations. This is indicated by using the suffix `_sl` (for "specimen level") in the simulation.

```{r}
library(paleoTS)
temp_res = 0.001                                              # temporal resolution in Myr
seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk_sl(sigma = 1, mu = 3, y0 = 0) |>                   # simulate trait evolution
  time_to_strat(adm, destructive = FALSE) |> 
   reduce_to_paleoTS() |># transform into stratigraphic domain
  plot(xlab = paste0("Stratigraphic position [",get_L_unit(adm),"]"), # plot
       ylab = "Trait value") 
```

see `?reduce_to_paleoTS` for details. We can then directly pass the results to the model selection options available in paleoTS:

```{r}
temp_res = 0.01                                              # temporal resolution in Myr
seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk_sl(sigma = 1, mu = 3, y0 = 0) |>                   # simulate trait evolution
  time_to_strat(adm, destructive = FALSE) |> 
   reduce_to_paleoTS() |>
  fit4models()
```

## Tasks

You can also solve the first two tasks interactively using [DarwinCAT](https://utrecht-university.shinyapps.io/DarwinCAT/).

::: callout-important
### Task 1: Trait evolution in the time domain

Plot the different modes of evolution in the time domain.

What is their internal variation due to randomness (i.e., how much do results differ if you plot them multiple times)?\
What is the biological meaning of the model parameters?
:::

::: callout-important
### Task 2: Stratigraphic effects on trait evolution

Transform the different modes of evolution into the stratigraphic domain and plot them.

Which modes are most/least affected by stratigraphic biases?\
Do you see any systematic changes?
:::

::: callout-important
### Task 3: Model selection in the stratigraphic domain

Run model selection on your simulations using the `paleoTS` package, and compare results from the time and the stratigraphic domain.

Do you see any systematic biases?
:::

## Solutions

You can also explore this topic interactively using [DarwinCAT](https://utrecht-university.shinyapps.io/DarwinCAT/).

::: {.callout-note collapse="true"}
### Solution Task 1: Trait evolution in the time domain

1.  Plot the different modes of evolution in the time domain.

    -   What is their internal variation due to randomness (i.e., how much do results differ if you plot them multiple times)?\
    -   What is the biological meaning of the model parameters?

```{r}
min_t = 0       # beginning observation [Myr]
max_t = 2       # end observation [Myr]
t_step = 0.01   # temporal resolution [Myr]
mean = 1        # model parameter, to modify
sd = 1          # model parameter, to modify
seq(from = min_t, to = max_t, by = t_step) |>
  stasis(mean = mean, sd = sd) |>
  plot(type = "l",
       xlab = "Time [Myr]",
       ylab = "Trait value",
       main = "Trait evolution: Stasis")
```

For stasis, the `mean` parameter determines the mean trait value, and the `sd` (standard deviation) parameter determines the spread of trait values. Trait values at individual times are uncorrelated, so if the temporal resolution is increased, trait evolution appears more irregular.

```{r}
min_t = 0      # beginning observation [Myr]
max_t = 2      # end observation [Myr]
t_step = 0.01  # temporal resolution [Myr]
sigma = 1      # model parameter, to modify
mu = 1         # model parameter, to modify
y0 = 1         # model parameter, to modify
seq(from = min_t, to = max_t, by = t_step) |>
  random_walk(sigma = sigma, mu = mu, y0 = y0) |>
  plot(type = "l",
       xlab = "Time [Myr]",
       ylab = "Trait value",
       main = "Trait evolution: Random walk")
```

For the random walk model, the `y0` parameter determines the initial trait value. The parameter `mu` determines the directionality, with positive (negative) values leading to increasing (decreasing) trait values with time. The `sigma` parameter determines the amount of randomness in the system, with larger values reflecting more randomness. Internal variability is high for larger values of `sigma`, meaning individual lineages simulated will differ a lot from each other.\
Important endmembers are: `sigma` set to 0, trait evolution is purely deterministic and specified by `mu` and `y0`. If `mu` and `y0` are set to 0, trait evolution follows an unbiased random walk model. If `sigma` is set to 1, the model corresponds to a Brownian motion. For low values of `sigma` and `mu`, the random walk can be mistaken for stasis.

```{r}
min_t = 0      # beginning observation [Myr]
max_t = 2      # end observation [Myr]
t_step = 0.01  # temporal resolution [Myr]
theta = 1      # model parameter, to modify
sigma = 0.1    # model parameter, to modify
mu = 0         # model parameter, to modify
y0 = 1         # model parameter, to modify
seq(from = min_t, to = max_t, by = t_step) |>
  ornstein_uhlenbeck(mu = mu, theta = theta, sigma = sigma, y0 = y0) |>
  plot(type = "l",
       xlab = "Time [Myr]",
       ylab = "Trait value",
       main = "Trait evolution: Ornstein-Uhlenbeck process")
```

For the Ornstein-Uhlenbeck (OU) model, the parameter `y0` represents the initial trait value. The parameter `mu` determines the long term mean of the process, `theta` how fast this mean is approached, and `sigma` how strong the influence of randomness is. This model is used to model convergence from an initial trait value `y0` to an optimum `mu` under pressure of selection `theta` and random influences `sigma`.\
Internal variation is determined by the `sigma` parameter, where larger values result in greater variation in simulated trait values.\
Important endmembers of the OU model are:

-   autocorrelated stasis, when `y0 = mu` (start in the optimum)

-   directional evolution when the optimum `mu` is far away from the initial trait value `y0`, and selection is present

-   random walk when selection `theta` is 0
:::

::: {.callout-note collapse="true"}
### Solution Task 2: Stratigraphic effects on trait evolution

2.  Transform the different modes of evolution into the stratigraphic domain and plot them.
    -   Which modes are most/least affected by stratigraphic biases?\
    -   Do you see any systematic changes?

Example code for the random walk model is provided. This can be expanded by replacing the random walk by stasis or the Ornstein-Uhlenbeck process.

```{r}

adm_list = list()                                     # storage for age-depth models
dist = paste(seq(2, 12, by = 2), "km", sep = "")      # distances from shore
for (pos in dist){                                    # construct ADMs for all distances
  adm_list[[pos]] = tp_to_adm(t = scenarioA$t_myr,        
                             h = scenarioA$h_m[,pos],
                             T_unit = "Myr",
                             L_unit = "m")
}

pos = "2km"              # position on platform
mu = 2                   # model parameters
sigma = 1                # model parameters
y0 = 0                   # model paramters
adm = adm_list[[pos]]    # select age-depth model
t_step = 0.001           # temporal resolution [Myr]
## plot trait evolution in the time domain
seq(from  = min_time(adm), to = max_time(adm), by = t_step) |>
  random_walk(sigma = sigma, mu = mu, y0 = y0) |>
  time_to_strat(adm, destructive = FALSE) |>
  plot(ylab = "Trait value",
       type = "l",
       orientation = "lr",
       xlab = "Height [m]",
       main = paste0("Trait evolution ", pos, " from shore") )
```

Stasis is unaffected by stratigraphic biases. This is because if there is no net change of traits in the time domain, losing parts of the lineage because of hiatuses does not affect the overall pattern.

For the random walk and the Ornstein-Uhlenbeck process, the degree of bias depends on the directionality of evolution. The more the traits change over a hiatus or a condensed interval, the more the stratigraphic expression of trait evolution differs from the one in the time domain. As a result, stratigraphic biases are strongest in the random walk model for large `mu`, and in the Ornstein-Uhlenbeck model for large differences between `y0` and `mu`.

The effects are strongest if

1.  There are long hiatuses or

2.  There are long intervals of sedimentary condensation.

As a result, stratigraphic biases are driven by maximum hiatus duration on the platform top, and driven by sedimentary condensation on the slope (see task on age-depth models). This effect is more pronounced the more directional evolution is.
:::

::: {.callout-note collapse="true"}
### Solution Task 3: Model selection in the stratigraphic domain

TBA

```{r}
temp_res = 0.1                                              # temporal resolution in Myr
seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk_sl(sigma = 1, mu = 3, y0 = 0) |>                   # simulate trait evolution
  time_to_strat(adm, destructive = FALSE) |> 
   reduce_to_paleoTS() |>
  fit9models()
```
:::
