---
title: "Trait Evolution"
format: html
editor: visual
bibliography: refs/refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(15487)
```

## Trait evolution in stratigraphic context

This practical explores how trait evolution is influenced by stratigraphic architectures.

## Introduction

Fossils found in a section can provide important insight into evolutionary dynamics over timescales not accessible to direct observations in the presents. The morphological evolution of these fossils allows us to construct *stratophenetic series*, which are successions of fossils and their traits observed within a stratigraphic section [@dzik]. Such fossil time or stratigraphic series allow us to examine how the mode of evolution changes over macroevolutionary timescales, and what the dominant mode of evolution is.

There are multiple "modes of evolution" discussed in the literature, such as stasis, random walk, or Ornstein-Uhlenbeck, which represents convergence to a phenotypic optimum. Each mode reflects an idea of the driving forces of evolution, e.g., the random walk model assumes change in traits is cumulative and random.

::: callout-tip
## Experience the modes of evolution

See our interactive app DarwinCAT at <https://utrecht-university.shinyapps.io/DarwinCAT/> and the associated [Open Educational Resource](https://zenodo.org/records/15667235) if you want to use it for teaching yourself.
:::

Because of the incompleteness of the fossil record, evolutionary biologists tend to be generally skeptical of inferences made from fossil time series, which is most famously reflected in the gradualism vs. punctuated equilibrium debate [@eldredge1972]. However, it seems that current methods of reconstructing the mode of evolution are surprisingly robust with respect to records with regular gaps [@hohmann_identification_2024].

In this practical, we will build on our understanding of age-depth models and stratigraphic architectures to explore how the stratigraphic record shapes our understanding of morphological evolution along a lineage. Specifically, we will examine when and how stratigraphy can lead to erroneous identification of the mode of evolution.

## Coding cheat sheet

### Modes of evolution in the time domain

There are three modes of evolution implemented in the `StratPal` package: `stasis`, `random_walk`, and `ornstein_uhlenbeck`. You can plot them using piping via

```{r models_evol_time_domain}
library(StratPal)
library(admtools)
pos = "2km"
adm = tp_to_adm(t = scenarioA$t_myr,        
                h = scenarioA$h_m[,pos],
                T_unit = "Myr",
                L_unit = "m")
temp_res = 0.001                                              # temporal resolution in Myr
seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk(sigma = 1, mu = 0, y0 = 0) |>                   # simulate trait evolution
  plot(xlab = paste0("Time [",get_T_unit(adm),"]"),           # plot
       ylab = "Trait value",
       type = "l") 
  

```

To plot other modes, replace `random_walk` by one of the other modes (`stasis` or `ornstein_uhlenbeck`). If you use stasis, remove the unused arguments - this mode of evolution is simulated using fewer parameters than the others.

### Transforming trait evolution into the stratigraphic domain

To transform trait evolution into the stratigraphic domain, insert `time_to_strat` with the age-depth model of your choice into your pipeline:

```{r models_evol_strat_domain}
temp_res = 0.001                                              # temporal resolution in Myr
seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk(sigma = 1, mu = 3, y0 = 0) |>                   # simulate trait evolution
  time_to_strat(adm, destructive = FALSE) |>                  # transform into stratigraphic domain
  plot(ylab = paste0("Stratigraphic position [",get_L_unit(adm),"]"), # plot
       xlab = "Trait value",
       type = "l",
       orientation = "du") 
```

This will transform your simulations from the time domain into the stratigraphic domain. We use the plotting option `orientation = "du"` to plot stratigraphic height on the y axis. Use `"lr"` to plot it on the x axis.

### Model selection using the paleoTS package

To work with model selection, we need to simulate traits of individuals, not just means of populations as we did above. This is indicated by using the suffix `_sl` (for "specimen level") in the simulation.

```{r paleoTS_sim}
library(paleoTS) # load paleoTS package for model selection & plotting
temp_res = 0.1                                             # temporal resolution in Myr
seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk_sl(sigma = 1, mu = 3, y0 = 0, intrapop_var = 1, n_per_sample = 5) |>                   # simulate trait evolution
  time_to_strat(adm, destructive = FALSE) |> 
  reduce_to_paleoTS() |># transform into stratigraphic domain
  plot(xlab = paste0("Stratigraphic position [",get_L_unit(adm),"]"), # plot
       ylab = "Trait value") 
```

We use `reduce_to_paleoTS` to convert the specimen level simulations from `StratPal` to the sample level summaries used by `paleoTS`, see `?reduce_to_paleoTS` for details. We can then directly pass the results to the model selection options available in paleoTS (`fit3models`, `fit4models` and `fit9models`, see `paleoTS` for details):

```{r paleoTS_model_selection}
temp_res = 0.1                                              # temporal resolution in Myr
intrapop_var = 1 # trait variance per sampling location/ fossil assemblage
n_per_sample = 5 # number of fossils found per sampling location / fossil assemblage

seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk_sl(sigma = 1, mu = 3, y0 = 0, intrapop_var = intrapop_var, n_per_sample = n_per_sample) |>                   # simulate trait evolution
  time_to_strat(adm, destructive = FALSE) |> 
  reduce_to_paleoTS() |>
  fit9models(pool = FALSE)
```

See the help of the `paleoTS` package for details on the output and `fitSimple` to fit more specialized models; more information can be found in [@hunt2006].

## Tasks

You can solve the first two tasks interactively using [DarwinCAT](https://utrecht-university.shinyapps.io/DarwinCAT/).

::: callout-important
### Task 2.1: Trait evolution in the time domain

Plot the three modes of evolution in the time domain, varying the parameters.

What is their internal variation due to randomness (i.e., how much do results differ if you plot them multiple times)?\
What is the biological meaning of the model parameters?
:::

::: callout-important
### Task 2.2: Stratigraphic effects on trait evolution

Transform the modes of evolution you simulated in the previous step into the stratigraphic domain and plot them.

Which modes are the most/least affected by stratigraphic biases?\
Do you see any systematic changes?
:::

::: callout-important
### Task 2.3: Model selection in the stratigraphic domain

Run model selection on your simulations using the `paleoTS` package, and compare results from the time and the stratigraphic domain.

Do you see any systematic biases?\
Do the results match how you would describe the mode of evolution?
:::

## Solutions

You can also explore this topic interactively using [DarwinCAT](https://utrecht-university.shinyapps.io/DarwinCAT/).

::: {.callout-note collapse="true"}
### Solution Task 2.1: Trait evolution in the time domain

Plot the different modes of evolution in the time domain.

What is their internal variation due to randomness (i.e., how much do results differ if you plot them multiple times)?\
What is the biological meaning of the model parameters?

```{r task_2_1_solution_stasis}
# Stasis
min_t = 0       # beginning observation [Myr]
max_t = 2       # end observation [Myr]
t_step = 0.01   # temporal resolution [Myr]
mean = 1        # model parameter, to modify
sd = 1          # model parameter, to modify
seq(from = min_t, to = max_t, by = t_step) |>
  stasis(mean = mean, sd = sd) |>
  plot(type = "l",
       xlab = "Time [Myr]",
       ylab = "Trait value",
       main = "Trait evolution: Stasis")
```

For stasis, the `mean` parameter determines the mean trait value, and the `sd` (standard deviation) parameter determines the spread of trait values. Trait values at individual times are uncorrelated, so if the temporal resolution is increased, trait evolution appears more irregular.

```{r task_2_1_solution_random_walk}
# Random walk
min_t = 0      # beginning observation [Myr]
max_t = 2      # end observation [Myr]
t_step = 0.01  # temporal resolution [Myr]
sigma = 1      # model parameter, to modify
mu = 1         # model parameter, to modify
y0 = 1         # model parameter, to modify
seq(from = min_t, to = max_t, by = t_step) |>
  random_walk(sigma = sigma, mu = mu, y0 = y0) |>
  plot(type = "l",
       xlab = "Time [Myr]",
       ylab = "Trait value",
       main = "Trait evolution: Random walk")
```

For the random walk model, the `y0` parameter determines the initial trait value. The parameter `mu` determines the directionality, with positive (negative) values leading to increasing (decreasing) trait values with time. The `sigma` parameter determines the amount of randomness in the system, with larger values reflecting more randomness. Internal variability is high for larger values of `sigma`, meaning individual lineages simulated will differ a lot from each other.\
Important endmembers are: `sigma` set to 0, trait evolution is purely deterministic and specified by `mu` and `y0`. If `mu` and `y0` are set to 0, trait evolution follows an unbiased random walk model. If `sigma` is set to 1, the model corresponds to a Brownian motion. For low values of `sigma` and `mu`, the random walk can be mistaken for stasis.

```{r task_2_1_solution_OU}
# Ornstein-Uhlenbeck
min_t = 0      # beginning observation [Myr]
max_t = 2      # end observation [Myr]
t_step = 0.01  # temporal resolution [Myr]
theta = 1      # model parameter, to modify
sigma = 0.1    # model parameter, to modify
mu = 0         # model parameter, to modify
y0 = 1         # model parameter, to modify
seq(from = min_t, to = max_t, by = t_step) |>
  ornstein_uhlenbeck(mu = mu, theta = theta, sigma = sigma, y0 = y0) |>
  plot(type = "l",
       xlab = "Time [Myr]",
       ylab = "Trait value",
       main = "Trait evolution: Ornstein-Uhlenbeck process")
```

For the Ornstein-Uhlenbeck (OU) model, the parameter `y0` represents the initial trait value. The parameter `mu` determines the long term mean of the process, `theta` how fast this mean is approached, and `sigma` how strong the influence of randomness is. This model is used to model convergence from an initial trait value `y0` to an optimum `mu` under pressure of selection `theta` and random influences `sigma`.\
Internal variation is determined by the `sigma` parameter, where larger values result in greater variation in simulated trait values.\
Important endmembers of the OU model are:

-   autocorrelated stasis, when `y0 = mu` (start in the optimum)

-   directional evolution when the optimum `mu` is far away from the initial trait value `y0`, and selection is present

-   random walk when selection `theta` is 0
:::

::: {.callout-note collapse="true"}
### Solution Task 2.2: Stratigraphic effects on trait evolution

Transform the different modes of evolution into the stratigraphic domain and plot them.

Which modes are most/least affected by stratigraphic biases?\
Do you see any systematic changes?

Example code for the random walk model is provided. This can be expanded by replacing the random walk by stasis or the Ornstein-Uhlenbeck process.

```{r task_2_2_solution_strat_effects}

adm_list = list()                                     # storage for age-depth models
dist = paste(seq(2, 12, by = 2), "km", sep = "")      # distances from shore
for (pos in dist){                                    # construct ADMs for all distances
  adm_list[[pos]] = tp_to_adm(t = scenarioA$t_myr,        
                             h = scenarioA$h_m[,pos],
                             T_unit = "Myr",
                             L_unit = "m")
}

pos = "2km"              # position on platform
mu = 2                   # model parameters
sigma = 1                # model parameters
y0 = 0                   # model paramters
adm = adm_list[[pos]]    # select age-depth model
t_step = 0.001           # temporal resolution [Myr]
## plot trait evolution in the stratigraphic domain
seq(from  = min_time(adm), to = max_time(adm), by = t_step) |>
  random_walk(sigma = sigma, mu = mu, y0 = y0) |>
  time_to_strat(adm, destructive = FALSE) |>
  plot(ylab = "Trait value",
       type = "l",
       orientation = "lr",
       xlab = "Height [m]",
       main = paste0("Trait evolution ", pos, " from shore") )
```

Stasis is unaffected by stratigraphic biases. This is because if there is no net change of traits in the time domain, losing parts of the lineage because of hiatuses does not affect the overall pattern.

For the random walk and the Ornstein-Uhlenbeck process, the degree of bias depends on the directionality of evolution. The more the traits change over a hiatus or a condensed interval, the more the stratigraphic expression of trait evolution differs from the one in the time domain. As a result, stratigraphic biases are strongest in the random walk model for large `mu`, and in the Ornstein-Uhlenbeck model for large differences between `y0` and `mu`.

The effects are strongest if

1.  There are long hiatuses or

2.  There are long intervals of sedimentary condensation.

As a result, stratigraphic biases are driven by maximum hiatus duration on the platform top. This effect is more pronounced the more directional evolution is.
:::

::: {.callout-note collapse="true"}
### Solution Task 2.3: Model selection in the stratigraphic domain

Run model selection on your simulation using the `paleoTS` package, and compare results from the time and the stratigraphic domain.

Do you see any systematic biases?

We run model selection both in the stratigraphic and time domain, fitting 9 models to the fossil time series. The example code performs this analysis for simulations under the random walk model, which can be changed out to the stasis or Ornstein-Uhlenbeck model.

```{r task_2_3_solution_model_selection_time_domain}
# time domain model selection
temp_res = 0.1                                              # temporal resolution in Myr
intrapop_var = 1 # trait variance per sampling location/ fossil assemblage
n_per_sample = 5 # number of fossils found per sampling location / fossil assemblage
traits_time_domain = seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk_sl(sigma = 1, mu = 3, y0 = 0, intrapop_var = intrapop_var, n_per_sample = n_per_sample) |>                   # simulate trait evolution
  reduce_to_paleoTS() 
plot(traits_time_domain, xlab = "Time [Myr]")
traits_time_domain |> fit9models(pool = FALSE)
```

```{r task_2_3_solution_model_selection_strat_domain}
# stratigraphic domain model selection
temp_res = 0.1                                              # temporal resolution in Myr
traits_strat_domain = seq(min_time(adm), max_time(adm), by = temp_res) |>           # time steps where the trait evolution is simulated
  random_walk_sl(sigma = 1, mu = 3, y0 = 0, intrapop_var = intrapop_var, n_per_sample = n_per_sample) |>                   # simulate trait evolution
  time_to_strat(adm, destructive = FALSE) |> 
  reduce_to_paleoTS() 
plot(traits_strat_domain, xlab = "Stratigraphic Position [m]")
traits_strat_domain |> fit9models(pool = FALSE)
```

General observation are:

-   Results are sensitive to randomness inherent in the randomness of the simulation, e.g., occasionally a random walk simulation will be identified as stasis even in the abscence of biases. This effect is additionally sensitive to parameters such as sample size and intrapopulation variance (`intrapop_var` and `n_per_sample`), reducing the power of the procedure and potentially leading to "analytical stasis", where low statistical power favors the recovery of stasis [@hannisdal2006].
-   Recovery of the correct (meaning, simulated) mode of evolution is better in the time domain, whereas model selection in the stratigraphic domain favors the recognition of punctuated modes of evolution [@hohmann_identification_2024]. This effect is strongest in the on the platform top, and disappears on the proximal slope as stratigraphic biases disappear.
-   As in the visual comparison, stratigraphic effects are strongest the more directional the mode of evolution is, and disappear for stasis.
-   Stratophenetic time series will thus favor both the recogniton of stasis and punctuated modes of evolution.
:::
